# Functions in `Python`

```{r setup_skill_functions_python, include = FALSE}
knitr::opts_chunk$set(cache = FALSE, message = FALSE, warning = FALSE)
library(dplyr,  quietly = TRUE, warn.conflicts = FALSE)
library(DiagrammeR,  quietly = TRUE, warn.conflicts = FALSE)
```

We've learned how to use built-in functions to analyze data, but sometimes it's more helpful to (A) do the math by hand or (B) write your own function. Let's learn how!

<br>
<br>

## Getting Started {-}

### Packages {-}

We'll use base Python here. No extra packages needed.

<br>

## Coding your own function!

`Functions` are machines that do a specific calculation using an input to produce a specific output.

```{r flowchart, echo = FALSE, fig.height=1.5, out.width="100%"}
c("graph LR",
  "subgraph before",
    "i[input data <i>x</i>]",
  "end",
  "subgraph function",
    "c[calculation]",
  "end",
  "subgraph after",
    "o[output data <i>y</i>]",
  "end",
  # Edges
  "i --> c",
  "c --> o"
) %>%
  paste(collapse = "\n ") %>%
  DiagrammeR::mermaid()
```

Below, we'll write an example function, called `add(a, b)`. 

- This function takes two `numeric` values, `a` and `b`, as `inputs`, and adds them together. 

- Using `def`, we'll tell `Python` that our function contains two inputs, `a` and `b`. 

- The function can involve multiple operations inside it. But at the end, you need to print *one* final output, or put `return` before your output.

```{python}
# Make function
def add(a, b):
  # Compute and directly output
  return a + b

add(1, 2)
```

```{python}
# This also works
def add(a, b):
  # Assign output to a temporary object
  output = a + b
  # Return the temporary object 'output'
  return output

add(1, 2)
```

<br>
<br>

## Functions with Default Inputs

You can assign default input values. Below, by default, `b = 2`. If we supply a different `b`, the default gets overwritten.

```{python}
def add(a, b = 2):
  return a + b
```

Let's try it!

```{python}
# Only provide 'a'
add(1)
# Provide both
add(1, 2)
# Change 'b'
add(1, 3)
```

```{python}
# clear data
del add
```

## Conclusion {-}

Great! Let's go make some functions!

```{python, include=FALSE}
globals().clear()
```

