# Appendix A: Visualization

Looking for extra information on making the perfect visualization? See below for extra

## Learning Check 7 {.unnumbered #LC7}

**Question**
 
Sometimes, the names of categories won't fit well. We can try the following. Compare these two plots. What did we do?

```{r plot17, eval = FALSE}
library(ggplot2)
diamonds = ggplot2::diamonds

# Plot 1
gg1 = gg1 = ggplot(data = diamonds, mapping = aes(x = cut, y = price, 
                                      group = cut, fill = cut)) +
  geom_boxplot()

# Plot2
ggplot(data = diamonds, mapping = aes(x = cut, y = price, 
                                      group = cut, fill = cut)) +
  geom_boxplot() + 
  coord_flip()
```

 
<details><summary>**[View Answer!]**</summary>
  
We used `coord_flip()` to **flip** the coordinates of the `x` and `y` axis, which gives our `cut` labels more room!

</details>
  
---

<br>
<br>
<br>

---

## Learning Check 8 {.unnumbered #LC8}

**Question**

Sometimes, the legend doesn't fit well. We can try this:

What happens when you change `legend.position` from `"right"` to `"bottom"` to `"left"` to `"top"`?

```{r plot18, eval = FALSE}
library(ggplot2)
diamonds = ggplot2::diamonds

ggplot(data = diamonds, mapping = aes(x = cut, y = price,
  group = cut, fill = cut)) +
  geom_boxplot() + coord_flip() +
  theme(legend.position = "bottom")

ggplot(data = diamonds, mapping = aes(x = cut, y = price,
  group = cut, fill = cut)) +
  geom_boxplot() + coord_flip() +
  theme(legend.position = "right")

ggplot(data = diamonds, mapping = aes(x = cut, y = price, 
  group = cut, fill = cut)) +
  geom_boxplot() + coord_flip() +
  theme(legend.position = "left")

ggplot(data = diamonds, mapping = aes(x = cut, y = price, 
  group = cut, fill = cut)) +
  geom_boxplot() + coord_flip() +
  theme(legend.position = "top")
```

  
<details><summary>**[View Answer!]**</summary>
  
These four values for `legend.position` will relocate the `fill` legend (and any other legends) to be at the top, bottom, left, or right of the visual!

</details>
  
---


## Breaking Up a Visual {-}

Finally, we might want to break up our visual into multiple parts. We can use `facet_wrap` to do this, but how exactly does it work? Let's test it out in the **Learning Check** below.

---

## Learning Check 10 {.unnumbered #LC10}

**Question**

What changed in the code below, and what did it result in?

```{r plot20, eval = FALSE}
library(ggplot2)
diamonds = ggplot2::diamonds

ggplot(data = diamonds, mapping = aes(x = price, fill = cut)) +
  geom_histogram(color = "white") + 
  facet_wrap(~cut) + # must be categorical variable
  labs(x = "Price (USD)",
       y = "Frequency of Price (Count)",
       title = "US Diamond Sales")
```

<details><summary>**[View Answer!]**</summary>

This visual split up our histograms into separate panels (making it much more readable), and easier to compare distributions. We write `facet_wrap(~` before the variable name (eg. `cut`) to specify that we want to split up the data by the values of `cut`. This sorts our rows of data into 5 different piles (since there are 5 different categories in `cut`) and makes a panel out of each.

</details>
  
---