---
title: "System Reliability and Six Sigma in R"
author: "Timothy Fraser, PhD"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "Your online textbook for learning reliability and six sigma techniques in R! Made for Cornell University Course SYSEN 5300."
---

```{r setup, include=FALSE, message=FALSE, warning = FALSE}
# Set options
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE)
options(knitr.graphics.auto_pdf = TRUE)
# Load packages
library(tidyverse)
library(bookdown)
library(htmltools)

# Clear environment
rm(list = ls())

# Load all functions in the f directory
funs = dir("f", full.names = TRUE)
funs %>% purrr::map(~source(.))

# Update any static data for us.
update(images = TRUE)
i = images() # Load images metadata list
```

# Introduction {.unnumbered #introduction}

Your online textbook for learning reliability and six sigma techniques in R! These coding workshops were made for Cornell University Course SYSEN 5300. Follow along with RStudio.Cloud to learn to apply six sigma techniques in R!

```{r img-cover, out.width = "100%", echo=FALSE, results='asis', fig.cap=paste(i$img_cover$caption)}
knitr::include_graphics(path = i$img_cover$file)
```

# (PART) Part I: Coding for Six Sigma {-} 

```{r child='workshop_1.Rmd'}
```

```{r child='skill_ggplot.Rmd'}
```

```{r child='skill_fmea.Rmd'}
```

# (PART) Part II: Statistics & Probability {-} 

```{r child='workshop_2.Rmd'}
```

```{r child='workshop_3.Rmd'}
```

```{r child='workshop_4.Rmd'}
```

```{r child='workshop_5.Rmd'}
```

```{r child='workshop_6.Rmd'}
```

```{r child='workshop_7.Rmd'}
```

```{r child='workshop_9.Rmd'}
```

```{r child='workshop_10.Rmd'}
```

# (PART) Part III {-} 





# References

# (APPENDIX) Appendix {-} 